Scaffolder - Software for Microbial Genome Scaffolding.

### 2011-02-08 text/src/article.tex

% }}}
% {{{ Background
\section*{Background}
 Text for this section.\cite{koon,oreg,khar,zvai,xjon,schn,pond,smith,marg,hunn,advi,koha,mouse}
% }}}
% {{{ Implementation
%%%%%%%%%%%%%%%%%%%%%%
\section*{Implementation}
  Text for this section \ldots
% }}}
\section*{Results and Discussion} %{{{1

\subsection*{Scaffolder Simplifies Genome Finishing} %{{{2

The Scaffolder software allows manual or computational joining of multiple
discontiguous nucleotide sequences together into a contiguous super sequence.
Simple plain-text configuration files are used to select the nucleotide
sequences in the scaffold and how they should be joined together. This is
called the ``scaffold file''. \pb

The scaffold file is read by the scaffolder software to determine how the
scaffold super sequence is generated. In addition to specifying which
nucleotides should be used to generate the draft super sequence, the scaffold
file allows the sequences to trimmed down to smaller sub-sequences and to
reverse complement the sequence if necessary. \pb

There may be unknown regions in the sequences used in the scaffold. Region of
nucleotide sequence the assembler was unable to determine the nucleotide
sequence. The process of genome finishing may involve determining the correct
nucleotide sequence for these regions then filling the gaps with the additional
correct sequence. The scaffold file specifically allows these inserts to be
added to scaffold. These inserts can be treated the same way as a larger
contig: trimmed and/or reversed to match the required gap region. \pb

When scaffolding a genome there may be regions of sequence approximate size
region may be estimated from mate-pair insert size or from mapping the contigs
to reference genome. These unknown regions are useful to join non-contiguous
sequences together by the estimated size of the region whilst still
highlighting areas in the genome to be resolved. These unknown regions can be
specified in the scaffold file for the above reasons. \pb

The original raw nucleotide sequence for each assembled contig or insert
sequence is maintained in a fasta file. The fasta file of contigs is a common
output for assemblers and as such should be readily available after assembly
without additional processing. The choice of the nucleotide sequence to use in
the scaffold is specified using the first space-delimited word from the fasta
header of the sequence. By maintaining the nucleotide sequences in a separate
file this preserves the original raw sequence in the fasta file. Furthermore
this separates the actual nucleotide sequence from the determination of how it
is selected and edited in the final genome sequence. \pb

\subsection*{Defining a scaffold in a text file} %{{{2

The scaffold file is written using `YAML', a data format designed to be both
machine readable and manually editable. An example of a scaffold file in YAML
format is shown in Figure 1. This example scaffold file illustrates the
available features in scaffolder. The basic layout of the scaffold file is as
a list - a series of entries where each entry defines a region of sequence to
be generated for the final super sequence. In the final super sequence each of
the entries are joined together to generate a single continuous sequences. \pb

\subsubsection*{Simple sequence region} %{{{3

The first line of the scaffold file begins with three dashes. This is
a requirement to indicate the start of a YAML formatted document. Line 2 begins
with a dash character `-' denoting the first entry in the sequence scaffold.
Line 3 is indented by two spaces. Whitespace is used to delimite attributes in
the file. Lines with an incremented level of white space are considered linked
to the first preceeding non-indented line. \pb

The ``sequence'' tag used on line 3 indicates that this region is in the
scaffold will contain nucleotide sequence from the fasta file. The following
line indented by two spaces and indicating an attribute of the sequence entry
defined on line 3. In this case the ``source'' keyword identifies the name of
the sequence to be inserted into the scaffold. As described above the first
space-delimited word of the fasta header should be used to identify the
sequence. Taken together lines 2-4 describe the first entry in the scaffold is
a simple sequence which the identifier `sequence'. \pb

\subsubsection*{Unresolved sequence region} %{{{3

Line 5 has no whitespace indentation and begins with a dash and therefore
describes the next entry in the scaffold. This entry is identified by the
``unresolved'' tag, and specifies a region of unknown nucleotides but known
length. The following line specifies the size of this unknown region. Lines 5-6
therefore insert a region of 20 `N' characters in the super sequence directly
following the region defined by the first entry on lines 2-5. \pb

\subsubsection*{Trimmed sequence region with multiple inserts} %{{{3

The last entry adds the sequence 'sequence2' in the scaffold file. This entry
demonstrates three further attributes for a sequence entry.The `start' and
`stop' tags are used to trim the sequence used in the scaffold to these
coordinates inclusively. The ``reverse'' tag is used to reverse complement this
region in the final super sequence. \pb

This region further includes the ``inserts'' tag to update the inserted
sequence with additional regions of sequence. These inserts are also added as
a YAML list with each insert listed with a dash on one line followed by the
remainder of the insert errata. \pb

Lines 16-19 illustrate an insert using similar attributes to that of a sequence
entry. The reverse, start and stop tags trim the inserted sequence and the
`source' tag identifies the corresponding fasta sequence. The ``open'' and
``close'' tags on lines 20-21 determine where the insert is added to the
enclosing sequence.  These coordinates inclusively replace the region within
these coordinates with the specified fasta sequence. The open and close
positions need not be the same size as the inserted sequence. \pb

The next insert on lines 23-24 specifies only the open position. This
illustrate that only the one of the `open' or `close' tags is required.  When
only open or close tag is used then the opposing position is calculated from
the length on the insert fasta sequence. This allows inserts to partially fill
gap regions in sequences. \pb

\subsubsection*{Scaffolder file attribute processing} %{{{3

The sequence entry in the scaffold file allows multiple sequence-editting
attributes which when applied in different combinations will result in
different output sequence. An example is reverse complementing the sequence
then follwing adding sequence inserts. Performing these actions in the reverse
order will produce a difference output sequence. This section therefore
outlines the scaffolder protocol for processing sequence entries. \pb

\paragraph{Reverse sort inserts:} If the sequence entry contains a list of
inserts these are reverse sorted by close position. The insert with the
largest close position is first in the list and inserts with the smallest
close position is last. This preserves the original indented coordinates for
insert during the process of adding inserts to the sequence. \pb

\paragraph{Update sequence with insert:} Each insert sequence is processed
according to any specified attributes. The insert is added to the enclosing
sequence in the region specified by the open and close coordinates. \pb

If the insert added to the sequence is a different length than the region
replaced the sequence start and stop co-ordinates are updated. For example if
a 7bp insert replaced a 5bp region the stop coordinate of the enclosing
sequence would be incremented by the 2bp difference. \pb

\paragraph{Triming sequence:} If the sequence is has start or stop sequences
specified these are used to trip the sequence to these co-ordinates
inclusively. \pb

\paragraph{Reverse complement sequence:} If the ``reverse'' keyword is
specified the sequence is reverse complemented. This is final step in
processing the sequence with results in the sequence being returned. \pb

\subsection*{Scaffolder usage} %{{{2

Scaffolder provides a standardised set of Ruby classes and methods (the
application programming interface or API) allowing the scaffold to be
manipulated programmatically. This allows scaffolder to be integrated into
existing genomics workflows or used with Ruby build tools such as using Rake.
\pb

Scaffolder also provides a command line interface (CLI) through a scaffolder
binary. This the scaffold sequence to be produced and validated without
requiring any Ruby knowledge. The scaffolder CLI behaves as a standard Unix
tool - returning appropriate exit statuses and providing appropriate manual
pages. The use of both these scaffolder interfaces is outlined in the following
sections. \pb

\subsubsection*{Installation} %{{{3

Scaffolder relies on both the ruby programming language and the RubyGems
package management system to work. These are either already present easily
installed on recent distributions of Mac OSX or Linux. The presences of these
tools can be tested by typing the following on the command line. \pb

\begin{verbatim}
  ruby -v
  gem -v
\end{verbatim}

If both these commands return without error this indicates the necessary
system requirements to install and run scaffolder. The scaffolder API and CLI
can then be installed through RubyGems with the following command.

\begin{verbatim}
  gem install scaffolder scaffolder-tools
\end{verbatim}

To ensure that the scaffolder CLI is accessible the RubyGems `bin' directory
should be in the command line path. Assuming that RubyGems are installed in
~/.gem this can be done by adding the following to the ~/.bashrc file. This
should be adjusted for other shells such as zsh etc..

\begin{verbatim}
  export PATH=~/.gem/bin:$PATH
\end{verbatim}

\subsubsection*{Command Line Interface} %{{{3

The Scaffolder CLI provides functions to validate the scaffold file and build the draft scaffold sequence. Assuming scaffolder has been installed and the
binary is available in the command line path the following command should run without error and list the available functions. \pb

\begin{verbatim}
  scaffolder
\end{verbatim}

Help on a specific command can be found by typing `help' followed by the name
of the command. This will return the corresponding Unix manual page. The
following example will return the manual page for the sequence command. \pb

\begin{verbatim}
  scaffolder help sequence
\end{verbatim}

The ``sequence'' command builds the fasta sequence from the scaffold and fasta
file. Running this command requires the locations of both the scaffold file and
the fasta file of sequences. This will then return the complete scaffold
sequence. \pb

\begin{verbatim}
  scaffolder sequence scaffold.yml sequences.fna
\end{verbatim}

Scaffolder provides a function to test the scaffold for overlapping insert
coordinates. Overlapping inserts will result in unexpected behaviour as the
coordinates for the first insert may change the insert location for subsequent
inserts. The ``validate'' command will check the scaffold for any such inserts
and return them as a YAML list. \pb

\begin{verbatim}
  scaffolder validate scaffold.yml sequences.fna
\end{verbatim}

\subsubsection*{Application programming interface} %{{{3

Scaffolder provides an API allowing Ruby programs to manipulate the scaffold.
This API is detailed in full on the scaffolder website. The following code
snippet outlines how the API may be used to create a script to replicate the
function of the `scaffolder sequence' command described above. \pb

\begin{verbatim}
  #!/usr/bin/env ruby

  require 'rubygems'
  require 'scaffolder'

  scaffold_file = ARGV.shift
  sequence_file = ARGV.shift

  scaffold = Scaffolder.new(
    YAML.load(File.read(scaffold_file)),
    sequence_file)

  sequence = scaffold.map{|entry| entry.sequence }.join

  puts sequence
\end{verbatim}

This code snippet illustrates that the scaffold object is initialised with two
arguments: the YAML loaded scaffold and the location of the fasta sequence
file. The scaffold is an enumerable object as the next line illustrates the
trivial code required to generate the scaffold sequence. Further details of
the classes and method related to each scaffold entry can be found in the
documentation for the Scaffolder::Region class. \pb

\subsection*{Discussion of scaffolder approach} %{{{2

The main aims of scaffolder are two fold. The first is usability and the second is reproducibility.

\subsubsection{Simple to use}

Scaffolder is designed to be as simple to use as possible. Assuming the
existence of Ruby and RubyGems, scaffolder can be installed with a single
command. Scaffolder uses a minimal and compact syntax to described the
scaffold. This syntax is simple to construct and edit whilst being succinct and
readable in describing how the scaffold was constructed. Scaffolder provides
extensive documentation both for the command line tools and for the Scaffolder
Ruby API.

\subsubsection{Reproducible genome construction}

Constructing a genome sequence using a plain text scaffold file means the same
sequence is always output for the same file. The sequence produced from
manually joining separate contigs in a text editor can not be reliably
reproduced. Constructing the scaffold in a plain text file means that the
sequence is reproducible. Furthermore the source of each region in final
sequence can be identified. The scaffold file provides a readable record of
how the sequence was constructed. The scaffold file also means the scaffold is
easier to edit once already constructed. Much more difficult to do when working
on the sequence alone. The YAML text format is also much easier to compare
differences using standard diff commands. This makes scaffold files much more
simple to store in version control and to see the stages in the construction
of the genome sequence.

\section*{Conclusions} %{{{1

Scaffolder follows the Unix philosophy ``Do one thing and do it well''.
Scaffolder focuses on creating a simple to use and install tool for generating
a genome scaffold. The YAML data format for creating a scaffold file is
standardised and easily manipulated programmatically. This thereby follows
a second Unix tenet: ``If your data structures are good enough, the algorithm
to manipulate them should be trivial.''

\section*{Availability and Requirements} %{{{1

>>>>

% }}}
\section*{Background} %{{{1
%{{{2

High throughput sequencing can be reads short regions of a genome sequence to
produce hundreds of thousands to millions of lines of nucleotide sequence. The
current status of parallel nucleotide sequencing is limited to producing reads
less than $>$1000 nucleotides long. Therefore to obtain a contiguous complete
genome sequence these sequence reads must therefore be joined together. \pb

The process of joining sequencing reads together is described as the
`assembly' stage in a genome project. Assembly software joins these numerous
sequence reads together into contiguous regions (`contigs') composed of many
overlapping individual reads. The goal of the assembly stage is produce
a continuous genome sequence describing the complete genome from start to
finish. \pb

Often repetitive nucleotide `repeat' regions in the genome or biased and
incomplete sequencing data may prevent the genome being assembled into
a continuous sequence. This sequence data can however describe the majority of
the genetic information in the genome and in the case of microbes and moving
forward with the current sequence data at this stage may be the best
investment of time and money \cite{branscomb2002}. \pb

A continuous high-quality `finished' genome sequence provides a much greater
depth of information as compared to a draft sequence
\cite{parkhill2002,fraser2002}. Examples of depth of information are complete
resolution of repeat regions and the exact distances between genomic elements.
The process of moving from a draft sequence to a finished continuous sequence
involves the process of `scaffolding' the contigs produced by the assembly
software together. \pb

\subsection*{Scaffolding} %{{{2

Scaffolding can be described as the process of moving from a series of
disconnected contigs to complete continuous sequence of all the contigs joined
together. The scaffolding process may not be completed but may still resolve
and join some contigs together or add additional regions of sequence to fill
gaps. In the case of a microbial genome, the process of scaffolding may
include the following steps:

\subsubsection*{Contig Orientation} %{{{3

Sequence reads may be produced from either DNA strand during the sequencing
process. Therefore the resulting contigs produced from these sequences may are
likely to be representative from either of the genome strands. Contig
orientation involves correcting the sequence data so that all contigs
represent sequence data from the same DNA strand. In the case of microbes this
orientation is often the 5`-3` strand from the origin of DNA replication.

\subsubsection*{Contig Ordering} %{{{3

The source location of each contig from the genome is unknown. Contig ordering
represents determining the order of contigs which most accurately represent
the actual genome sequence. This process can also determine if any contigs
represent extra-genomic DNA such as plasmids and therefore should be separated
from the genomic contigs. Contigs are usually ordered in reference to the
origin of replication where the contig enclosing this region is the first
contigs and subsequent contigs are labelled in the direction of DNA
replication.

\subsubsection*{Contig Distancing} %{{{3

Once the order of contigs in the genome sequence is known it is useful to
determine their approximate distance from each other. This allows the size of
the gaps between contigs to be determined. Filling in these regions in with
unknown nucleotide characters allows a draft continuous representation of the
genome to determined.

\subsubsection*{Gap Closing and Filling} %{{{3

The final stage of producing a genome sequence is to close and fill gaps in
the sequence. This may require returning to the laboratory to perform
additional sequencing to close gaps or \emph{in silico} methods may be able to
close gaps using existing data. Once all gaps in a scaffold have been closed
the genome may be considered finished. Though additional sequencing may be
performed to `proof' any low coverage regions for erroneous nucleotide
insertions, deletions or substitutions.

\subsection*{Computational Methods for Scaffolding} %{{{2

Scaffolding may be considered a combination of \emph{in vitro} and \emph{in
silico} methods. Several steps may be performed computationally by using
available data to generate a scaffold. Examples may be the use of pair reads
sequencing, where reads a known distance apart, or from comparing contigs to
a reference genome. In the case where \emph{in silico} methods are unable to
complete the scaffold in may be necessary to use PCR to amplify then use
traditional Sanger sequencing to find the sequence and close gaps. Obviously
\emph{in silico} methods are preferable as they less costly in laboratory time
and materials \cite{nagaryan2010}. \pb

For these reason may examples exist for software to generate genome scaffolds
from assembly data. These software use a variety of approaches. Synteny
approaches use a reference genome to order and orient contigs
\cite{richter2007,zhao2008}. Repeat regions which may be responsible for many
of the gaps between contigs may be reassembled using algorithms specifically
for repeat resolution \cite{mulkov2002,koren2010}. Uniquely matching reads may
be used to try and bridge gaps between scaffolds \cite{tsai2010}. Whole genome
may use heuristic graph algorithms to build a path between contigs using
multiple input data \cite{pop2004,dayarian2010}. Finally when the scaffold
cannot be completely resolved \emph{in silico} software packages exist to
suggest the likely primers necessary for determining the sequence in gap
regions \cite{gordon2001,nagaryan2010}. Taken together these tools provide
a wide variety of possible options for scaffolding a genome should an
assembler be unable to produce a continuous sequence from the sequencing
reads. \pb

Scaffolding however will inevitably require manual editing if computational
methods do not completely resolve the final genome sequence. This involves
moving an joining larges blocks of nucleotide text together in what the genome
finisher perceives to be the correct scaffold. This process may be expected to
be error-prone. Further manual editing cannot be reliably reproduced which
precludes reproducibility in a genome scaffolding process. \pb

\subsection*{The Scaffolder Software} %{{{2

The software outlined here ``Scaffolder'' is aimed at joining contigs together
to build a genome scaffold. Scaffolder uses a specific data syntax to define
how contigs are joined together. Furthermore scaffolder allows inserts into
contigs and unknown regions in a scaffold to be defined. This syntax allows
a scaffold to be edited in much easier way than editing large blocks of
nucleotides. Further more this makes genome scaffolds reproducible.

\section*{Implementation} %{{{1

  Text for this section \ldots
% }}}
% {{{ Results and Discussion

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Results and Discussion %%
%%
\section*{Results and Discussion}
  \subsection*{Results sub-heading}
    \subsubsection*{This is a sub-sub-heading}
      Sub-sub-sub-headings are made with the \textsl{\\subsubsection} command. \pb
      pb at end of lines ensures correct paragraph spacing.\pb
	  Text for this sub-sub-section \ldots
    \subsubsection*{Another sub-sub-sub-heading}
      Text for this sub-sub-section \ldots

  \subsection*{Another results sub-heading}
    Text for this sub-section \ldots

  \subsection*{Yet another results sub-heading}
    Text for this sub-section.  More results \ldots
% }}}
% {{{ Conclusions
%%%%%%%%%%%%%%%%%%%%%%
\section*{Conclusions}
  Text for this section \ldots
% }}}
% {{{ Availability and Requirements
\section*{Availability and Requirements}
